---
title: "R Notebook"
output: html_notebook
---

# Mean sensitivity

This is straightforward. The MS is the mean of all sensitivity values (except blind spot).

```{r msens, eval=FALSE}

msens <- apply(vf, 1, mean)

```

\[\mathrm{MS} = \frac{1}{n}\sum_{i=1}^{n} S_i\]

# SD of sensitivity

Again, just standard deviation of all sensitivity values (except blind spot).

```{r ssens, eval=FALSE}

msens <- apply(vf, 1, mean)

```

\[\text{SS} = \sqrt{\frac{1}{n} \sum_{i=1}^{n} (S_i - \overline{S})^2}\]

# Mean deviation

For the HFA, this is a weighted mean of all deviation values (except blind spot). 

```{r tdev, eval=FALSE}

tmd   <- apply(td, 1, wtd.mean, weights = wtd / sum(wtd))

```

I did not understand, why the weights are normalized to 1 (=divided by their sum) only for tmd, but not for the other calculations. However, it seems to make no difference, because the formula for weighted calculation normalizes anyway.

```{R}

library(Hmisc)

values <- rnorm(10, mean = 5)
weights <- 1:10

wtd.mean(values, weights = weights)
wtd.mean(values, weights = weights/sum(weights))

```

\[\mathrm{MD} = \frac{\sum_{i=1}^{n} \frac{1}{Var(S_{normal})} \cdot TD_i}{\sum_{i=1}^{n} \frac{1}{Var(S_{normal})}}\]

# SD of deviation

```{r tsd, eval=FALSE}

tsd   <- sqrt(apply(td, 1, wtd.var, weights = wtd))

```

# Mean pattern deviation

This parameter is not used regularly, I think.

I'm not sure how PD is calculated in the package. I think that Iván wrote in the manuscript, that it is calculated by subtracting the general height from the TD values. I cannot find it in the code. 

In the PSD formula from the IPS guidelines, however, MD is subtracted from the TD.

In the latter case, I have the feeling that PMD would result in 0. If you subtract the mean of a bunch of numbers from these numbers, the average results in 0?

```{r pmd, eval=FALSE}

pmd   <- apply(pd, 1, wtd.mean, weights = wpd)

```

# Pattern standard deviation

```{r psd, eval=FALSE}

psd   <- sqrt(apply(pd, 1, wtd.var, weights = wpd))

```

# Visual field index

Should we document the calculation and the weights in detail. Iván, from your comments, it sounds like 

